---
title: "Week1-Exercises-Solutions"
format:
  html:
    theme: cosmo
    code-fold: true
    code-summary: "Show the code"
  pdf:
    documentclass: scrreprt
    keep-md: true
    keep-tex: true
---

```{r, setup, include=FALSE}

library(readxl)
library(ggplot2)
library(Statamarkdown)
stataexe <- "/Applications/Stata/StataBE.app/Contents/MacOS/StataBE"
knitr::opts_chunk$set(engine.path=list(stata=stataexe))
knitr::opts_knit$set(root.dir = '../Data') # Changes the working director to the Data folder
```

# Exercise solutions

### Week 1 {.unnumbered}

#### (a) and (b) {.unnumbered}

**Stata code and output**
```{stata, collectcode=TRUE, collapse=TRUE }
/* Part a */
use hers_subset
reg DBP BMI

# Part b
gen BMI5 = BMI / 5
reg DBP BMI5
```

**R code and output**
```{r, collapse = TRUE}
# Part a
hers_subset <- read.csv("hers_subset.csv")
lm.hers <- lm(DBP ~ BMI, data = hers_subset)
summary(lm.hers)
confint(lm.hers)

# Part b
hers_subset$BMI5 <- hers_subset$BMI / 5
lm.hers <- lm(DBP ~ BMI5, data = hers_subset)
summary(lm.hers)
```

We find evidence that diastolic blood pressure increases as body mass index increases (P = 0.029). For every one kg/m^-2^ increase in BMI, the mean diastolic blood pressure increases by 0.22mmHg, and we are 95% confident the true increase lies between 0.023 and 0.42mmHg. BMI accounts for 1.7% of the overall variability in diastolic blood pressure.


If a one kg/m^-2^ increase in BMI accounts for a 0.22mmHg increase in DBP, then a 5kg/m^-2^ increase in BMI accounts for a `5x0.22 = 1.1`mmHg increase in DBP. We can confirm this in Stata or R by creating a new covariate `BMI5` which is BMI scaled by a factor of 1/5 (so that a 1 increase in BMI5 corresponds to a 5 increase in BMI).


#### (c) {.unnumbered}

From 3.3.7 of the textbook, the standard error of the regression coefficient is as follows:

$$ \text{se}(\beta_1) = \frac{\text{Root mean squared error}}{\sigma_x \sqrt{(n-1)} } $$
We can use R or Stata to calculate $\sigma_x = 5.5879$. Subsituting this in we obtain $\text{se}(\beta_1) = 9.357/(5.5879 \sqrt{275}) = 0.101$ in agreement with the Stata and R output.

The t-value is the regression cofficient divided by it's standard error $t = 0.222 / 0.101 = 2.2$, and the P-value can be calculated by looking up a corresponding t-table with $n-2 = 276-2 = 274$ degrees of freedom:


Stata code and output
```{r, echo=1, collapse = TRUE, comment=""} 
## Stata code
x <- capture.output(stata("disp tprob(274,2.2)"))
cat("disp tprob(274,2.2)")
cat(paste("##", x[-1]), sep = "\n") 
```

R code and output
```{r, echo=TRUE, collapse = TRUE}
(1-pt(2.2,274))*2
```

R^-2^ is the fraction of the total variance explained by the model so is equal to $R^{-2} = 423.88/24412.77 = 0.017$. These two variances are default output in Stata. In R the model sum of squares and residual sum of squares can be obtained  using `anova(lm.hers)`, after which the R^{-2} can be calculated.

#### (d), (e) and (f) {.unnumbered}

Stata code and output
```{stata, collectcode=TRUE, collapse=TRUE }
use hers_subset
reg DBP BMI

lincom _cons + 28*BMI

set obs 277
replace BMI = 28 in 277
predict fitDBP
predict seprDBP, stdf
gen upper = fitDBP + 1.96*seprDBP in 277
gen lower = fitDBP -1.96*seprDBP in 277

list BMI fitDBP lower upper in 277
```

R code and output
```{r, collapse = TRUE}
hers_subset <- read.csv("hers_subset.csv")
lm.hers <- lm(DBP ~ BMI, data = hers_subset)

new_observation <- data.frame(BMI = 28)
predict(lm.hers, newdata = new_observation, interval="confidence")
predict(lm.hers, newdata = new_observation, interval="prediction")
```
We predict that the mean diastolic blood pressure for those with a BMI of 28kgm^-2^ to be 74mmHg. We are 95% confident the true mean lies between 72.9mmHg and 75.2mmHg. We expect that 95% of women with that BMI will have a diastolic blood pressure between 55.6mmHg and 92.5mmHg.


