---
title: "Week8-Exercises-Solutions"
format:
  pdf:
    documentclass: scrreprt
    keep-md: true
    keep-tex: true
editor_options: 
  chunk_output_type: inline
---

```{r, setup, include=FALSE}
library(ggplot2)
library(Statamarkdown)
stataexe <- "/Applications/Stata/StataBE.app/Contents/MacOS/StataBE"
knitr::opts_chunk$set(engine.path=list(stata=stataexe))
knitr::opts_knit$set(root.dir = '../Data') # Changes the working director to the Data folder
```

# Exercise solutions

Use the tools above to investigate the ideal number of knots for the week 7 investigation between HDL and BMI.


Below shows the code and output for running a cubic spline with 3, 4, and 5 knots. We see that AIC and BIC consistently increase with increasing number of knots, and that 3 knots seems to generate the lowest AIC and BIC values. Therefore 3 knots is the most appropriate function form for BMI in this data.

Stata code and output
```{stata, collectcode=TRUE, collapse=TRUE }
clear
use hersdata.dta

mkspline BMIsp3k = BMI, cubic nknots(3) 
regress HDL BMIsp3k1 BMIsp3k2 age nonwhite smoking drinkany
estimates stats

mkspline BMIsp4k = BMI, cubic nknots(4) 
regress HDL BMIsp4k1 BMIsp4k2 BMIsp4k3 age nonwhite smoking drinkany
estimates stats

mkspline BMIsp5k = BMI, cubic nknots(5) 
regress HDL BMIsp5k1 BMIsp5k2 BMIsp5k3 BMIsp5k4 age nonwhite smoking drinkany
estimates stats
```

R code and output.
```{r, collapse = TRUE}
require(haven)
library(rms)

hers<-read_dta("https://www.dropbox.com/s/ndtd4o20qogq7fv/hersdata.dta?dl=1") 
hers<-data.frame(hers)

# reduce the dataset and remove missing
hers1<-hers[,c("HDL", "BMI", "age", "nonwhite", "smoking", "drinkany")]
hers1<-na.omit(hers1)
dim(hers1) 
# 2745 after removing the missing

ddist <- datadist(hers1)
options(datadist='ddist')

# Run models with 3 - 7 knots
fit.3knots <-  lm(HDL ~ rcs(BMI,3) + age + nonwhite + smoking +drinkany, data = hers1)
fit.4knots <-  lm(HDL ~ rcs(BMI,4) + age + nonwhite + smoking +drinkany, data = hers1)
fit.5knots <-  lm(HDL ~ rcs(BMI,5) + age + nonwhite + smoking +drinkany, data = hers1)

AIC(fit.3knots)
AIC(fit.4knots)
AIC(fit.5knots)

BIC(fit.3knots)
BIC(fit.4knots)
BIC(fit.5knots)
```




































